\chapter{Introduction}

The CIPR in-house reservoir flow simulator models fluid flows in
porous media, such as oil, water and gas. This mass flow is modelled
using a generalized version of the empirical Darcy law. Energy
transfer is modelled using Darcy's law for the convective flux and
Fourier's law for the conductive flux. Principles of mass, energy, and
volume balance provide a set of coupled equations in molar masses,
temperature, and pressure.

Currently, the CIPR in-house simulator has the following features:
\begin{itemize}
\item An extensible compositional fluid system is used:
  \begin{itemize}
  \item The fluid consists of discrete chemical components.
  \item Phase properties may be calculated using either a black-oil or
    a compositional model.
  \item Temperature variations are accounted for using energy
    conservation.
  \end{itemize}
\item Control volume IMPEM discretisation:
  \begin{itemize}
  \item Pressure is calculated implicitly using a volume balance
    method.
  \item Advective fluxes for the masses and temperature are treated
    explicitly.
  \item Two- or multi-point flux approximations.
  \item Parallelized with a distributed memory solver.
  \end{itemize}
\item Simple and flexible grids:
  \begin{itemize}
  \item Cartesian structured meshes with local grid refinements, which
    the simulator generates.
  \item Unstructured meshes, imported from external meshing tools or
    from corner point grids.
  \end{itemize}
\end{itemize}

%----------------------------------------------------------------------

\csection{License}

Copyright 2004-2006 Centre for Integrated Petroleum Research (CIPR).
All rights reserved.

Redistribution and use in source and binary forms, with or without
modification, are permitted provided that the following conditions are
met:
\begin{enumerate}

\item Redistributions of source code must retain the above copyright
  notice, this list of conditions and the following disclaimer.

\item Redistributions in binary form must reproduce the above
  copyright notice, this list of conditions and the following
  disclaimer in the documentation and/or other materials provided with
  the distribution.

\item All advertising materials mentioning features or use of this
  software must display the following acknowledgement:

  \begin{quote}
    This product includes software developed by the\\
    Centre for Integrated Petroleum Research\\
    and its contributors.
  \end{quote}

\item Neither the name of the Centre nor the names of its contributors
  may be used to endorse or promote products derived from this
  software without specific prior written permission.

\end{enumerate}

\begin{sloppypar}
  THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
  "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
  LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
  A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
  OWNER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
  SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
  LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
  DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
  THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
  (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
  OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
\end{sloppypar}
