\chapter{Notation}

The physical units of an entity is given in brackets following its
description. These units are all SI. Dimensionless units are given the
unit $[-]$.

%----------------------------------------------------------------------

\section{Indices}

Superscripts refer to phases, while subscripts refer to components
(mass or energy).

\begin{description}
\item[$\ell$] Phase superscript: water ($\ell=w$), oil ($\ell=o$), and
  gas ($\ell=g$).
\item[$\nu,\mu$] Component subscripts. $0$ is taken to be water.
\end{description}

%----------------------------------------------------------------------

\section{Volumes}

\begin{description}
\item[$V$] Bulk volume. $\left[\meter\cubed\right]$
\item[$V^p$] Pore volume, $V^p=\phi V$. $\left[\meter\cubed\right]$
\item[$V^f$] Fluid volume, $V^f=\sum_\ell V^\ell$. $\left[\meter\cubed\right]$
\item[$V^\ell$] Phase volume. $\left[\meter\cubed\right]$
\item[$R$] Residual volume, $R=V^p-V^f$. $\left[\meter\cubed\right]$
\item[$\phi$] Porosity, the ratio between pore volume and bulk volume:
  $\phi=V^p/V$. $\left[-\right]$
\item[$V^r$] Rock volume, $V^r=V - V^p$. $\left[\meter\cubed\right]$
\item[$c^r$] Rock compressibility. $\left[1\per\pascal\right]$
\item[$\phi_0$] Reference (initial) porosity. $\left[-\right]$
\item[$p_0$] Reference (initial) pressure. $\left[\pascal\right]$
\item[$S^\ell$] Phase saturation, $S^\ell=V^\ell/V^f$.
  $\left[-\right]$
\item[$z_D$] Datum depth. $\left[\meter\right]$
\item[$p_D$] Pressure at datum depth. $\left[\pascal\right]$
\end{description}

%----------------------------------------------------------------------

\section{Mass flow}

\begin{description}
\item[$N_\nu$] Moles of component $\nu$. $\left[\mole\right]$
\item[$\xi^\ell$] Molar density of phase $\ell$.
  $\left[\mole\per\meter\cubed\right]$
\item[$C_\nu^\ell$] Molar fraction of component $\nu$ in phase $\ell$.
  $\left[-\right]$
\item[$C_\nu$] Molar fraction of component $\nu$.  $\left[-\right]$
\item[$N_\nu^\ell$] Moles of component $\nu$ in phase $\ell$.
  $\left[\mole\right]$
\item[$\vec u^\ell$] Darcy phase velocity of phase $\ell$.
  $\left[\meter\per\second\right]$
\item[$\lambda^\ell$] Mobility of phase $\ell$,
  $\lambda^\ell=k_r^\ell/\mu^\ell$.
  $\left[1\per\pascal\usk\second\right]$
\item[$k_r^\ell$] Relative permeability. $\left[-\right]$
\item[$\mu^\ell$] Phase viscosity. $\left[\pascal\usk\second\right]$
\item[$\tensor K$] Absolute permeability. $\left[\meter\squared\right]$
\item[$\Psi^\ell$] Phase potential. $\left[\pascal\right]$
\item[$p^\ell$] Phase pressure. $\left[\pascal\right]$
\item[$\gamma^\ell$] Specific phase gravity, $\gamma^\ell=g\rho^\ell$.
  $\left[\pascal\per\meter\right]$
\item[$g$] Gravitational acceleration constant. $g=9.81
  \meter\per\second\squared$
\item[$\rho^\ell$] Phase density.
  $\left[\kilogram\per\meter\cubed\right]$
\item[$z$] Coordinate in the upwards (anti-gravity) direction.
  $\left[\meter\right]$
\item[$p_c^{ow}$] Oil/water capillary pressure. $\left[\pascal\right]$
\item[$p_c^{go}$] Gas/oil capillary pressure. $\left[\pascal\right]$
\item[$q_\nu$] Source of component $\nu$. $\left[\mole\per\second\right]$
\item[$q^\ell$] Source of phase $\ell$. $\left[\mole\per\second\right]$
\item[$q_T$] Total source/sink. $\left[\mole\per\second\right]$
\item[$f^\ell$] Fractional flow function,
  $f^\ell=\lambda^\ell/\sum_\ell \lambda^\ell$. $\left[-\right]$
\end{description}

%----------------------------------------------------------------------

\section{Energy flow}

\begin{description}
\item[$\rho^r$] Rock density. $\left[\kilogram\per\meter\cubed\right]$
\item[$e^r$] Rock energy density. $\left[\joule\per\kilogram\right]$
\item[$e^\ell$] Phase energy density. $\left[\joule\per\kilogram\right]$
\item[$h^\ell$] Phase enthalpy, $h^\ell=e^\ell+p^\ell/\rho^\ell$.
  $\left[\joule\per\kilogram\right]$
\item[$\kappa$] Rock thermal conductivity.
  $\left[\watt\per\meter\usk\kelvin\right]$
\item[$T$] Temperature. $\left[\kelvin\right]$
\item[$q^e$] Energy source. $\left[\watt\right]$
\item[$\delta^T$] Heat capacity. $\left[\joule\per\kelvin\right]$
\end{description}

%----------------------------------------------------------------------

\section{Thermodynamics}

\begin{description}
\item[$M_\nu$] Molecular weight of component $\nu$.
  $\left[\kilogram\per\mole\right]$
\item[$\mathcal{G}$] Set of components grouped to the gas
  pseudo-component.
\item[$\mathcal{O}$] Set of components grouped to the oil
  pseudo-component.
\item[$N_\text{HC}$] Moles of hydrocarbons. $\left[\mole\right]$
\item[$Z_o$] Fraction of oil in the hydrocarbon system,
  $Z_o=N_o/N_\text{HC}$. $\left[-\right]$
\item[$Z_g$] Fraction of gas in the hydrocarbon system,
  $Z_g=N_g/N_\text{HC}$. $\left[-\right]$
\item[$C_{g,B}^o$] Bubble-point curve. $\left[-\right]$
\item[$C_{o,D}^g$] Dew-point curve. $\left[-\right]$
\item[$L$] Liquid hydrocarbon fraction, $L=N^o/N_\text{HC}$. $\left[-\right]$
\item[$V$] Liquid hydrocarbon fraction, $V=N^g/N_\text{HC}$. $\left[-\right]$
\item[$\bar h^\ell$] Phase molar enthalpy.
  $\left[\joule\per\mole\right]$
\end{description}

%----------------------------------------------------------------------

\section{Control volume discretisation}

\begin{description}
\item[$\Omega_i$] Control volume $i$.
\item[$\partial\Omega_i$] Surface of the control volume $\Omega$.
\item[$\partial\Omega_{i,j}$] Face $j$ of the surface
  $\partial\Omega_i$. $\partial\Omega_{i,j}$ is planar, meaning that
  the whole surface has just one normal vector $\vec n_{i,j}$.
  $\partial\Omega_i$ has no single normal vector.
\item[$\vec n_{i,j}$] Outward normal vector on $\partial\Omega_{i,j}$.
  $\left[-\right]$
\item[$\mathcal{M}_{i,k}^\tensor{K}$] Set of control volumes
  contributing to the Darcy flux across $\partial\Omega_{i,k}$.
\item[$\mathcal{M}_{i,k}^\kappa$] Set of control volumes contributing
  to the Fourier flux across $\partial\Omega_{i,k}$.
\item[$t_{ij}$] Coupling coefficient or transmissibility for the
  absolute permeability.
\item[$k_{ij}$] Coupling coefficient or transmissibility for the
  heat conductivity.
\item[$D_{i,k}^\ell$] Volumetric phase Darcy flux in the outward
  normal direction on $\partial\Omega_{i,j}$.
  $\left[\meter\cubed\per\second\right]$
\item[$F_{i,k}^\ell$] Volumetric Fourier flux in the outward normal
  direction on $\partial\Omega_{i,j}$.
  $\left[\meter\cubed\per\second\right]$
\item[$r_i$] Residual in control volume $i$.
\item[$L_{i,j}$] Distance between control volume $i$ and $j$.
\end{description}

